Efficient Computation of Extended Sources

### 2021-05-24 repro/papers/sseprecond/paper.tex
interim

I will fully describe the preconditioner construction for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, the spatial positions of traces extracted by $P$ lie
on a plane, and the positions at which the extended
source $\bff$ is nonzero lie on another, parallel, plane. The data and
sources are further limited to pressure traces and constitutive law
defects (``pressure sources''). While this
``crosswell'' configuration simplifies the analysis underlying the
construction of approximate solutions for the source subproblem
\ref{eqn:esis}, it is only one of many transmission configurations for
which similar developments are possible. Perhaps the most important
alternative example is the diving wave configuration, which plays a
central role in contemporary FWI. I will discuss the generalization of
surface source extention to diving wave inversion at the end of the
paper.

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. These are essentially the same as the ``hyperbolic
Dirichlet-to-Neumann'' map that plays a prominent role in
thermoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. The obvious computation of these maps
- prescribe the pressure, solve the wave equation with this boundary
condition, read off the equivalent source - suffers from intrinsic
numerical inaccuracy. I suggest an alternative computationally
feasible approach.

@@

\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

@@

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, {\em provided} that an accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
An approximate inverse for $S^++_{z_s,z_r}$ would permit solution of
this equation for $\Lambda^+_{z_s}f_s$.

In this section I will explain how to build the approximate inverse.
The key ingredient in this {\em time reversal} construction is a
{\em local energy decay} assumption. Define the local energy in the
acoustic field $(p,\bv)$ at
time $t$ in a region $\Omega \subset \bR^3$ by limiting the integration
in the definition of total energy \ref{eqn:defae0} to $\Omega$:
\begin{equation}
  \label{eqn:defaeloc}
  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
\end{equation}
The necessary assumption is: local energy decay in the half-space
$\Omega = \{(x,y,z): z \le z_r\}$. The local energy in $\Omega$ of the
causal solution $(p^+,\bv^+)$ of system \ref{eqn:awepm} cannot decay,
since the source on $z=z_s<z_r$ radiates in both positive and negative
$z$ in forward time. That obstacle is resolved by a redefinition of
the ``downgoing'' concept. I will call  downgoing if there is a
source-free acoustic field $(\tilde{p}^+,\tilde{\bv}^+)$ in all of space-time for which
\begin{itemize}
\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
  the slab $\{(x,y,z,t): z_s \le z \le z_r\}$, and
\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
  as $t \rightarrow \infty$.
\end{itemize}
Conditions for local energy decay have been widely studied in various
settings, for instance obstacle scattering
\cite[]{MorawetzPhillipsLax:05}. \cite{Pauen2000NonTr-6079 gives a
comprehensive overview of this subject. For smooth (non-scattering)
acoustic parameters, as is assumed in this paper, the crux of the
matter is that rays carrying high-frequence energy exit the half-space
$\Omega$ in finite time. For the half-space
geometries considered here, local energy does not decay in general,
because of the existence of ``trapped'' rays traveling parallel to the
boundary and remaining interior to $\Omega$ for all time. If an
acoustic field is downgoing in the usual sense, that is, its
high-frequency energy is concentrated on rays that exit $\Omega$
within some common maxiumum time, then the results cited above show
that local energy in $\Omega$ decays (exponentially for $n=3$).

Starting from this observation, it is possible to show that if the
source vector $(h_s,f_s)^T$ has the property termed ``downgoing'' in
preceding sections, that is, emanates high-frequency energy only along
rays that leave $\Omega$ within a common maximum time, then the
solution $(p^+,\bv^+)$ of \ref{eqn:awepm} is downgoing in the sense
defined here. All of our numerical examples behave as one would expect
for downgoing fields.

An immediate consequence of the downgoing assumption stated above is that in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. For any choice of $T$, define
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
\end{eqnarray}
As will be shown later, under the downgoing assumption, $P_r p^+
\approx P_r \tilde{p}^+$.
Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
differs negligibly from the solution of
\begin{eqnarray}
\label{eqn:deltaawe}
  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
                                                               \nabla
                                                               \cdot
                                                               \delta \bv, \nonumber \\
  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
  P_r\delta p & = & 0,
\end{eqnarray}
in $\Omega \times \bR$. Since
$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
E$ in this field,
\[
  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
 \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
\]
satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
$P_s$ is not continuous, but it does so imply for purely downgoing
fields, as will be demonstrated later, so
\begin{equation}
  \label{eqn:approxinv0}
  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
\end{equation}

@@

\section{Fields, Traces and Sources}
In this section, $(p^{\pm}_0,\bv^{\pm}_0)$ denotes an acoustic field,
continuous in the region $z_s \le z $, obeying the homogeneous wave
equation in the interior $z_s < z $:
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_0}{\partial t} & = & - \nabla \cdot \bv^{\pm}_0, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_0}{\partial t} & = & - \nabla
                                                    p^{\pm}_0, \nonumber \\
  p^{\pm}_0 & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
  \bv^{\pm}_0 & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}
I will show how to view $(p^{\pm}_0,\bv^{\pm}_0)$ as the $z>z_s$ part
of a solution of the system \ref{eqn:awepm} in two different ways,
corresponding to two different extensions to the whole space. An
important relation between the components of the operator
${\cal S}^{\pm}_{z_s,z_r}$ follows (equation \ref{eqn:lamnull} below).

Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
as the solution in the two half-spaces $z <z_s, z>z_s$ of
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
\end{eqnarray}
By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solves the
inhomogeneous system
\begin{eqnarray}
\label{eqn:sawe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

This is precisely the causal option of system \ref{eqn:awepm} for
$f_s=0$ and $h_s=\Lambda^{\pm}_{z_s}\phi$. From \ref{eqn:awe1} and the definition
\ref{eqn:fwd},
\begin{equation}
  \label{eqn:lamfwd}
  \phi = S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s}\phi,
\end{equation}
that is,
\begin{equation}
  \label{eqn:lamfwd1}
  S^{\pm}_{z_s,z_s} = (\Lambda^{\pm}_{z_s})^{-1}.
\end{equation}

Similarly, suppose $\psi$ is a function on $z=z_s$, and that
$(p^{\pm}_2,\bv^{\pm}_2)$ solves
\begin{eqnarray}
\label{eqn:awe2}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = & - \nabla \cdot \bv^{\pm}_2, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & - \nabla p^{\pm}_2,\nonumber \\
  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sv^{\pm}_{2,z} &=& \psi.
\end{eqnarray}
in the two half-spaces $z<z_s, z > z_s$. In this case, $v^{\pm}_{2,z}$
is continuous at $z=z_s$ by construction. Viewed as a
distribution on all of $\bR^4$, $(p^{\pm}_2,\bv^{\pm}_2)$ solve the
system \ref{eqn:awepm} with $h_s=0$, and the discontinuity in pressure
across $z=z_s$ as the inhomogeneous term in Newton's law:
\begin{eqnarray}
\label{eqn:sawe2}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = &
                                                        - \nabla \cdot \bv^{\pm}_2, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & 
                                                    [p^{\pm}_2]|_{z=z_s}{\bf e_z}\delta(z-z_s) - \nabla p^{\pm}_2,\nonumber \\
  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}
This relation defines a map from the velocity boundary data $\psi$ to
the source in the velocity equation, having units of force/volume,
similar to $\Lambda^{\pm}_{z_s}$. As it's not needed in the sequel, I
won't name this map.

A closer relation between traces and sources follows from the
geometric optics analysis of the half-space solution, presented
below. I will show that, for propagating wave data,
\begin{equation}
  \label{eqn:odd1}
  \lim_{z\rightarrow z_s^+} v_{1,z} \approx - \lim_{z\rightarrow z_s^-}
  v_{1,z}
\end{equation}
Similarly,
\begin{equation}
  \label{eqn:odd2}
  \lim_{z\rightarrow z_s^+} p_{2} \approx - \lim_{z\rightarrow z_s^-}
  p_{2}
\end{equation}
for the velocity field solving \ref{eqn:awe1}, or equivalently
\ref{eqn:awe1}. Here ``$\approx$'' means in the sense of high
frequency asymptotics, that is, that the difference between the two
sides is relatively smooth, hence small if the data is highly
oscillatory. Similarly, for the pressure field solving \ref{eqn:awe2}
or \ref{eqn:sawe2}, the jump at $z=z_s$ is twice the limiting value
(trace) from above, up to a smooth error. Therefore
\begin{eqnarray}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s} &\approx& 2
                              \lim_{z\rightarrow z_s}v^{\pm}_{1,z},\\  \label{eqn:tracejump1}
  [p^{\pm}_2]|_{z=z_s} &\approx& 2 \lim_{z\rightarrow z_s^+}p^{\pm}_{2},\\  \label{eqn:tracejump2}
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%
These apparently harmless relations have a profound consequence,
because of a final step: identify the boundary data $\phi$ and $\psi$,
so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
that is, set
\begin{eqnarray}
  \phi & = & P_s p^{\pm}_0, \nonumber \\
  \psi & = & P_s v^{\pm}_{0,z}.
             \label{eqn:bv0}
\end{eqnarray}
Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
the same, and both are pressure components of solutions of the same homogeneous
acoustic system in $z>z_s$, the standard energy identity implies that
$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
$z>z_s$.

Therefore the assumption \ref{eqn:bv0} implies that
\begin{equation}
  \label{eqn:jump0}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]_{z=z_s} \approx 2P_s v^{\pm}_{0,z}  = 2\psi,
\end{equation}
and
\begin{equation}
  \label{eqn:jump1}
[p_2^{\pm}]_{z=z_s} \approx  2P_s p^{\pm}_{0}= 2 \phi.           
\end{equation}
In the notation of the last section, set $2 \phi = 2P_sp^{\pm}_0= f_s,
2\psi = 2P_s v^{\pm}_{0,z}= h_s$. Then relation \ref{eqn:jump0}
implies that
\begin{equation}
  \label{eqn:lamfh}
  \Lambda^{\pm}_{z_s}f_s \approx h_s.
\end{equation}

From the definitions in the last section and the system \ref{eqn:sawe1},
\begin{equation}
  \label{eqn:pident0}
P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
  S}^{\pm}\Pi_0^Th_s,
\end{equation}
and
\begin{equation}
  \label{eqn:pident1}
P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
\end{equation}
Similarly, from the relations above and \ref{eqn:sawe2},
\begin{equation}
  \label{eqn:vident0}
  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
\end{equation}
and
\begin{equation}
  \label{eqn:vident1}
  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
\end{equation}
Together with \ref{eqn:lamfh}, these equations imply that
\begin{equation}
  \label{eqn:lamnull}
{\cal  S}_{z_s,z_r}^{\pm}\Pi_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}\Pi_1^T.
\end{equation}
Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
in the sense that its composition with the block operator
$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.

\section{Parametrix via Time Reversal}

\section{Approximate Unitarity}
Recall that the fields $(p^{\pm},\bv^{\pm})$ in $z>z_s$ can be viewed
as either the solution of the system \ref{eqn:awepm} with pressure source $h_s=0$ and
a suitable choice of velocity source $f_s$, or of the same system with
suitable choice of $h_s$ and $f_s=0$, and that these choices are
related by equation \ref{eqn:lamfh}: $\Lambda^{\pm}f_s = h_s$. The
pressure traces at $z=z_r>z_s$ are the same: from the definitions
\ref{eqn:sdef}, \ref{eqn:vdef},
\begin{eqnarray}
  \label{eqn:sandv}
  S^{\pm}_{z_s,z_r}h_s& =& P_rp^{\pm},\nonumber \\
  V^{\pm}_{z_s,z_r}f_s &=& P_rv_z^{\pm}.
\end{eqnarray}
Applying the same reasoning to traces on $z=z_r$: create pressure and
velocity sources $h_r$ and $f_r$ at $z=z_r$ by odd extension of the
velocity, respectively pressure, fields, and using the jumps as source
coefficients. Then as before
$\Lambda^{\pm}_{z_r}f_r=h_r=-P_rv^{\pm}_z$ (the sign change is due to
$Lambda$ producing the jump, whereas $P_rv^{\pm}_z$ is the limit from
the left of the field $v^{\pm}_{1,z}$ of system \ref{eqn:awe1}), hence
$\Lambda^{\pm}_{z_r}P_rp^{\pm} = P_rv_z^{\pm}$. Combining this
observation with equations
\ref{eqn:lamfh}, \ref{eqn:sandv}, obtain (with roles of $z_s,z_r$ interchanged)
\[
V^{\pm}_{z_r,z_w}(-2P_rp^{\pm}) =
  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2P_rv_z^{\pm})  =
  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2\Lambda^{\pm}_{z_r}P_rp^{\pm})
\]
whence
\begin{equation}
  \label{eqn:preunit}
 V^{\pm}_{z_r,z_w}
 =\Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}\Lambda^{\pm}_{z_r}.
\end{equation}

In view of the relations \ref{eqn:trtrcomp}, equation
\ref{eqn:preunit} implies that
\[
  V^-_{z_r,z_s}=-(V^+_{z_s,z_r})^T = -\Lambda^{-}_{z_s}(S^+_{z_s,z_r})^T\Lambda^{-}_{z_r}.
\]
\[
  =-(\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T.
\]
Therefore the first equation in display \ref{eqn:approxinv} can be re-written
\[
  -4V^-_{z_r,z_s}S^+_{z_s,z_r} =
\]
\begin{equation}
  \label{eqn:baseunit}
 4 (\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T
  S^+_{z_s,z_r} \approx I
\end{equation}

\section{Derivations}
\subsection{Extension to whole space}
If $\phi$ is smooth and vanishes for large negative $t$, then
$(p_1,\bv_1)$ can be constructed by pasting together the smooth
solutions of the inhomogeneous Dirichlet problems in the half-spaces
$\pm z > z_s$. Construct these by extending $\phi$ smoothly to
$\pm z>z_s$, and solve systems similar to \ref{eqn:awe1} for the
difference with $(p_1,\bv_1)$ by quoting the main results from
\cite{BlazekStolkSymes:13}. Since the right hand side in this
difference system is smooth, and the system is causal, the solution is
smooth and vanishes on $z=z_s$. Adding back the extension of $\phi$
produces smooth solutions in $\pm z > z_s$; define $(p_1,\bv)_1$ to be
these smooth functions in $z \ne z_s$. Since the boundary values for
pressure are the same, $p_1$ extends continuously to $z=z_s$. 
The pressure boundary value $\phi$ uniquely determines the solutions
in the two half-spaces, by standard energy estimates,

\subsection{Adjoint Computation}
The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
the adjoint state method, in this case a by-product of the
conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
  p^- + f_r \bf{e_z}) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
\end{equation}

The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
\begin{equation}
  \label{eqn:trsadj}
  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that

\section{Surface source-to-pressure operator}

>>>>

I will fully describe a preconditioner for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, the spatial positions of traces extracted by $P$ lie
on a plane, and the positions at which the extended
source $\bff$ is nonzero lie on another, parallel, plane. The data and
sources are further limited to pressure traces and constitutive law
defects (``pressure sources''). While this
``crosswell'' configuration simplifies the analysis underlying the
construction of approximate solutions for the source subproblem
\ref{eqn:esis}, it is only one of many transmission configurations for
which similar developments are possible. Perhaps the most important
alternative example is the diving wave configuration, which plays a
central role in contemporary FWI. I will discuss the generalization of
surface source extention to diving wave inversion at the end of the
paper.

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. These are essentially the same as the ``hyperbolic
Dirichlet-to-Neumann'' map that plays a prominent role in
thermoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang
  YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.

The obvious computation of the pressure-to-source map
- prescribe the pressure, solve the wave equation with this boundary
condition, read off the equivalent source - suffers from intrinsic
numerical inaccuracy. I suggest an alternative computationally
feasible approach.

@@

\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dsrcvzghh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

@@

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, {\em provided} that an accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
An approximate inverse for $S^++_{z_s,z_r}$ would permit (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$.

Recall that the source vector $(h_s,f_s)$ is assumed to produce a
downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
rays that make an angle with the vertical bounded below by a common
minimum angle. Such rays leave $\Omega$ within a common maximum time.

An immediate consequence of the downgoing assumption stated above is that in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. Indeed, for any choice of $T$, define
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega \times (-\infty,T]$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
\end{eqnarray}
That is, $\tilde{p}^-$ has the same boundary value as
$p^+$ for $t<T$, but $(\tilde{p}^-,\tilde{\bv}^-)$ vanishes at
$t=T$. Suppose that $T$ is chosen so that the rays carrying
significant energy in $(p^+,\bv^+)$ have left the half-space $\Omega$
at time $T$. Then

As will be shown later, under the downgoing assumption, $P_r p^+
\approx P_r \tilde{p}^+$.
Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
differs negligibly from the solution of
\begin{eqnarray}
\label{eqn:deltaawe}
  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
                                                               \nabla
                                                               \cdot
                                                               \delta \bv, \nonumber \\
  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
  P_r\delta p & = & 0,
\end{eqnarray}
in $\Omega \times \bR$. Since
$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
E$ in this field,
\[
  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
 \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
\]
satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
$P_s$ is not continuous, but it does so imply for purely downgoing
fields, as will be demonstrated later, so
\begin{equation}
  \label{eqn:approxinv0}
  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
\end{equation}

@@
